Longitudinal Principal Manifold Estimation

### 2023-10-02 paper/main.tex
Updates from Overleaf
overleaf updates
fix table formatting
readd main.tex
Ani revisions

\title{Longitudinal Principal Manifold Estimation}
\author[1]{Robert Zielinski}
\author[2]{Kun Meng}
\author[1]{Ani Eloyan}
\affil[1]{Department of Biostatistics, Brown University}
\affil[2]{Division of Applied Mathematics, Brown University}

@@

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}). Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}). For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes. Image segmentation allows for the extraction of subcortical structures from MRI images of the brain for further analysis.

Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. In studies analyzing a large number of images, this approach may impose prohibitive costs. Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}). Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}). While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of four years in the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice. In a larger comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies].

@@

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003a}), are ill-equipped to reach meaningful estimates across multiple time points. This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time.

Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to estimate the appropriate principal manifold at each given time point. We then smooth over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. The regularization involved in this process mitigates the effects of variability between study visits.

Some previous work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by \cite{kinsonLongitudinalPrincipalComponent2020}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to to incorporate longitudinal information when reducing neuroimaging information to lower dimensional spaces, but estimated parameterizations for baseline images and longitudinal image differences separately, where we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. To our knowledge, the approach to longitudinal smoothing in the general-purpose manifold learning setting taken in this article is novel.

The remainder of this article is laid out as follows: In Section 2, we review the principal manifold framework proposed in \cite{mengPrincipalManifoldEstimation2021}. Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this. Section 4 demonstrates the performance of this approach on simulated data. In Section 5, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease. The article concludes with a discussion of the method's contributions in Section 6.

\section{Principal Manifolds and Estimation Algorithm}

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of the data. However, the PS framework in \cite{hastiePrincipalCurves1989} is flawed by the model complexity selection, which \cite{duchamp1996extremal} pointed out. Motivated by the penalization approaches in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework and algorithm. Since our LPME is based on the PME, we review the PME in this section for the audience's convenience. The following concepts and notations are necessary for both the PME and our proposed LPME, and they will be used throughout this paper
\begin{itemize}
\item $d$ and $D$ are two positive integers denoting the dimensions of the low-dimensional manifold and high-dimensional space, respectively; 
\item for any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$;
    \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$, where $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ denotes the collection of continuous vector-valued functions mapping from $\mathbb{R}^d$ to $\mathbb{R}^D$; 
    \item for every $f\in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, the image of $f$ is referred to as the $d$-dimensional manifold $M_f^d$ determined by $f$, i.e., $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$;
    \item For each $f\in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and each point $x\in\mathbb{R}^D$, $\pi_f(x)$ is a point in $\mathbb{R}^d$ and denotes the ``projection index" of $x$; roughly, $\pi_f(x)$ is a point in $d$-dimensional space such that $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$; $\pi_f(x)$ is a generalization of the counterpart in \cite{hastiePrincipalCurves1989}, and its details can be found in Section 2 of \cite{mengPrincipalManifoldEstimation2021};
    \item suppose $u(r_1,\ldots,r_d)$ is a scalar-valued function defined on $\mathbb{R}^d$; $\nabla^{\otimes 2}u=\left(\frac{\partial^2 u}{\partial r_i \partial r_j}\right)_{1\le i,j\le d}$ denotes the Hessian matrix of $u$, and $\|\nabla^{\otimes 2}u\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2} u(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2\right)^{\frac{1}{2}}$; in addition, $\|\nabla^{\otimes 2} u\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2dr $, and 
    \begin{align*}
        \nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\},
    \end{align*}
    where $\mathcal{D}'$ denotes the collection of distributions (also known as generalized functions, see \cite{duchon1977splines} and Chapter 6 of \cite{rudin1991functional});
    \item $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$ denotes the collection of vector-valued functions $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ whose components are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d})$; besides, $\mathcal{C}_\infty\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)=\mathcal{C}_\infty(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$; the following functional of $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ measures the curvature of the manifold $M_f^d$ and will be used throughout this paper
    \begin{equation}
  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \|\nabla^{\otimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2  = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
\end{equation}
\end{itemize}
With the preparations above, the definition of principal manifolds proposed in \cite{mengPrincipalManifoldEstimation2021} is presented as follows
\commentout{

@@

To completely define principal manifolds, \cite{mengPrincipalManifoldEstimation2021} let $\|Pf\|_{\mathcal{H}} = \|f''(r)\|_{\mathbb{R}^{D}}^2$, which provides a greater penalization of roughness than using the first derivative. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
\begin{equation}
  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
\end{equation}
referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:}
\begin{definition}
  \label{def:principal_manifolds}
  Let $\boldsymbol{X}$ be a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments. Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
  \begin{align}\label{eq:5}
&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f). \nonumber
  \end{align}
  A manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $\boldsymbol{X}$ with the tuning parameter $\lambda$ if 
  \begin{equation}\label{eq:7}
    \begin{aligned}
        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \\ 
    &\text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}. 
    \end{aligned}
  \end{equation}
\end{definition}
In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, the first segment of the functional $\mathcal{K}_{\lambda, \mathbb{P}}$ provides a measure of the expected distance between the provided $D$-vector $\mathbf{X}$ and the value of $\mathbf{X}$ when projected onto the manifold. Meanwhile the second segment of $\mathcal{K}_{\lambda, \mathbb{P}}$ penalizes the roughness of the function $f$, with the magnitude of the penalty dependent on the provided value of $\lambda$.

To estimate principal manifolds under this definition, \cite{mengPrincipalManifoldEstimation2021} propose the Principal Manifold Estimation (PME) algorithm. This algorithm has three main steps: data reduction, fitting, and tuning (see Figure 2 of \cite{mengPrincipalManifoldEstimation2021}). \meng{I would suggest presenting Figure 2 of \cite{mengPrincipalManifoldEstimation2021} in the present paper.}
\begin{enumerate}
    \item In the data reduction step, the high-dimensional mixture density estimation (HDMDE) algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$. The number $N$ of clusters is chosen using an iterative hypothesis testing approach, and $N$ tends to be much smaller than $I$ (see the simulation study in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). 
    \item For each given value of $\lambda$, the fitting stage of the PME algorithm uses this reduced data set $\left\{\mu_j\right\}_{j=1}^{N}$ to find estimated function 
\begin{equation}
  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f), \label{eq:8}
\end{equation}
where $\mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2$ approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. 
\item The tuning stage chooses the optimal value of the parameter $\lambda$ by $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
\begin{equation}
  \mathcal{D}\left(\hat{f}_\lambda\right) = \frac{1}{I}\sum_{i=1}^{I}\left\|x_i - \hat{f}_\lambda\left(\pi_{\hat{f}_\lambda}(x_i)\right)\right\|_{\mathbb{R}^{D}}^2 \label{eq:10}
\end{equation}
is an empirical estimate of the MSD using the full dataset.
\end{enumerate}

Using results in \cite{duchon1977splines} and \cite{wahba1990}, \cite{mengPrincipalManifoldEstimation2021} show that the optimal function will be of a spline form. Specifically, when $d = 1$, the function will be a cubic smoothing spline, while the function is a thin plate spline when $d = 2$. Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in \cite{mengPrincipalManifoldEstimation2021}.

\meng{I commented out many literature review discussions about \cite{hastiePrincipalCurves1989} and \cite{smolaRegularizedPrincipalManifolds2001} to make this section more concise. In addition, if the formulas above are not referred to in our discussions, they shouldn't be given labels.}

\section{Longitudinal Principal Manifold Estimation Algorithm}

@@

\section{Simulations}

@@

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.222 (0.255) & 0.512 (0.409) & N/A & 30.5 (89.0) & 0.975 (0.770) & 1.43 (6.13) & N/A & N/A & N/A \\
    LPME & 0.125 (0.161) & 0.384 (0.659) & N/A & 27.8 (266) & 0.788 (0.843) & 1.18 (5.71) & N/A & N/A & N/A \\
    PME & 0.267 (0.862) & 0.839 (1.93) & N/A & 30.5 (89.2) & 1.03 (1.07) & 1.47 (6.14) & N/A & N/A & N/A \\
    PC/PS & 0.188 (0.244) & 0.599 (0.297) & N/A & 30.4 (89.0) & 0.930 (0.713) & 1.00 (2.14) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Mean (SD).}
  \label{table:simulation_results_mean}
\end{table}

>>>>

\title{Longitudinal Principal Manifold Estimation}
\author[1]{Robert Zielinski}
\author[2]{Kun Meng}
\author[1]{Ani Eloyan}
\author[*]{for the Alzheimer’s Disease Neuroimaging Initiative}
\affil[1]{Department of Biostatistics, Brown University}
\affil[2]{Division of Applied Mathematics, Brown University}

@@

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}). Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}). For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes. Image segmentation allows for the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. In studies analyzing a large number of images, this approach may impose prohibitive costs. Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}). Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}). While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies]. In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}). To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.

@@

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. This is because these methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits. Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, but estimated parameterizations for baseline images and longitudinal image differences separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

The remainder of this article is laid out as follows: In Section \ref{s:PME}, we review the principal manifold framework. In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus in individuals with Alzheimer's disease (AD). The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

\section{Principal Manifolds and Estimation Algorithm}\label{s:PME}

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection. In addition, applying the methods in high-dimensional settings requires is problematic. Motivated by the penalization approaches presented by in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, i.e. $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. 
\begin{itemize}
%    \item  = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$, where  denotes the collection of continuous vector-valued functions mapping from $\mathbb{R}^d$ to $\mathbb{R}^D$;

    \item suppose $u(r_1,\ldots,r_d)$ is a scalar-valued function defined on $\mathbb{R}^d$; $\nabla^{\otimes 2}u=\left(\frac{\partial^2 u}{\partial r_i \partial r_j}\right)_{1\le i,j\le d}$ denotes the Hessian matrix of $u$, and $\|\nabla^{\otimes 2}u\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2} u(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2\right)^{\frac{1}{2}}$; in addition, $\|\nabla^{\otimes 2} u\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2dr $, and 
    \begin{align*}
        \nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\},
    \end{align*}
    where $\mathcal{D}'$ denotes the collection of distributions (also known as generalized functions, see \cite{duchon1977splines} and Chapter 6 of \cite{rudin1991functional});
    \item $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$ denotes the collection of vector-valued functions $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ whose components are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d})$; besides, $\mathcal{C}_\infty\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)=\mathcal{C}_\infty(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$; the following functional of $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ measures the curvature of the manifold $M_f^d$ and will be used throughout this paper
    \begin{equation}
  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \|\nabla^{\otimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2  = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
\end{equation}
\end{itemize}
With the preparations above, the definition of principal manifolds proposed in \cite{mengPrincipalManifoldEstimation2021} is presented as follows

@@

To completely define principal manifolds, \cite{mengPrincipalManifoldEstimation2021} let $\|Pf\|_{\mathcal{H}} = \|f''(r)\|_{\mathbb{R}^{D}}^2$, which provides a greater penalization of roughness than using the first derivative. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
\begin{equation}
  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
\end{equation}
referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:

\begin{definition}
  \label{def:principal_manifolds}
  Let $\boldsymbol{X}$ be a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments. Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
  \begin{align}\label{eq:5}
&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f). \nonumber
  \end{align}
  A manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $\boldsymbol{X}$ with the tuning parameter $\lambda$ if 
  \begin{equation}\label{eq:7}
    \begin{aligned}
        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \\ 
    &\text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}. 
    \end{aligned}
  \end{equation}
\end{definition}
In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, the first segment of the functional $\mathcal{K}_{\lambda, \mathbb{P}}$ provides a measure of the expected distance between the provided $D$-vector $\mathbf{X}$ and the value of $\mathbf{X}$ when projected onto the manifold. Meanwhile the second segment of $\mathcal{K}_{\lambda, \mathbb{P}}$ penalizes the roughness of the function $f$, with the magnitude of the penalty dependent on the provided value of $\lambda$.

\section{Longitudinal Principal Manifold Estimation Algorithm}\label{s:LPME}

@@

\section{Simulations}\label{s:simulations}

@@

\begin{table}[h]
\footnotesize
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.222 (0.255) & 0.512 (0.409) & N/A & 30.5 (89.0) & 0.975 (0.770) & 1.43 (6.13) & N/A & N/A & N/A \\
    LPME & 0.125 (0.161) & 0.384 (0.659) & N/A & 27.8 (266) & 0.788 (0.843) & 1.18 (5.71) & N/A & N/A & N/A \\
    PME & 0.267 (0.862) & 0.839 (1.93) & N/A & 30.5 (89.2) & 1.03 (1.07) & 1.47 (6.14) & N/A & N/A & N/A \\
    PC/PS & 0.188 (0.244) & 0.599 (0.297) & N/A & 30.4 (89.0) & 0.930 (0.713) & 1.00 (2.14) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Mean (SD).}
  \label{table:simulation_results_mean}
\end{table}
